The socialization of meritocracy and market justice preferences at school

### 2024-08-05 MDPI-submit/template-mdpi.tex
Actualización paper anonimizado en .tex y orden de repo

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The socialization of meritocracy and market justice preferences at school},
  pdfauthor={Juan Carlos Castillo; Mauricio Salgado; Kevin Carrasco; Andreas Laffert},
  pdfkeywords={market
justice, meritocracy, socialization, family, schools, Chile},
  pdfcreator={LaTeX via pandoc},
  unicode=true,
	bookmarksopen={true},
	pdffitwindow=true, 
	colorlinks=true, 
	linkcolor=bluecite, 
	citecolor=bluecite, 
	urlcolor=bluecite, 
	hyperfootnotes=true, 
	pdfstartview={FitH},
	pdfpagemode= UseNone}

@@

\Author{Juan Carlos Castillo$^{1}$\orcidA{}, 
Mauricio Salgado$^{2}$\orcidB{}, 
Kevin Carrasco$^{3}$\orcidC{}~and~Andreas Laffert$^{4}$\orcidD{}
}
%\longauthorlist{yes}

% MDPI internal command: Authors, for metadata in PDF
\AuthorNames{
  Juan Carlos Castillo,
  Mauricio Salgado,
  Kevin Carrasco,
  Andreas Laffert}

% MDPI internal command: Authors, for citation in the left column
%\AuthorCitation{Lastname, F.; Lastname, F.; Lastname, F.}
%\usepackage{xstring}
\newcommand\getfirst[1]{\getfirstaux#1\relax} \def\getfirstaux#1#2\relax{#1}
\newcommand{\authorformatted}{Castillo, \getfirst{Juan Carlos}.; Salgado, \getfirst{Mauricio}.; Carrasco, \getfirst{Kevin}.; Laffert, \getfirst{Andreas}.}
\AuthorCitation{\authorformatted}

@@

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address {%
$^{1}$ \quad Department of Sociology, Universidad de Chile\\
$^{2}$ \quad CEP, Centro de Estudios Públicos\\
$^{3}$ \quad COES, Center for Social Conflict and Cohesion Studies\\
$^{4}$ \quad Institute of Sociology, Pontificia Universidad Católica de Chile\\
}

@@

% Contact information of the corresponding author
%% see: https://quarto.org/docs/journals/authors.html
\corres{Correspondence: juancastillov@uchile.cl}

@@

Robert E. Lane proposed the underpinnings of the concept of market
justice, which he differentiated from political justice. For him, ``it
is the genius of the market to stimulate wants without at the same time
stimulating a sense of deserving more than one gets''
\citeyearpar[p.~384]{lane_market_1986}. Contrary to the evidence that
unequal distribution produces feelings of dissatisfaction, anger, and
resentment that might motivate forms of collective action
\citep{greitemeyer_subjective_2016, mishra_subjective_2015, smith_relative_2012, power_deprivationprotest_2018},
Lane pointed out that in market settings, social comparisons are more
likely to motivate increased effort rather than feelings of acute
injustice because individuals attribute outcomes to their actions. In
this sense, unequal levels of well-being would be, to some extent, a
function of their talents and efforts, instead of being based on
distributive principles that characterize welfare states, such as need
and equality (see \citep{wilson_role_2003}).

@@

Our research aimed to examine the relationship between the perception of
meritocracy and the justification of market justice preferences among
eighth-grade students in Chile. From our knowledge this is the first
study addressing market justice preferences at school level, finding
that students exhibit large preferences for market justice (about one
third agree or strongly agree with them), a share that is quite larger
when compared with evidence in adult population. Perception of
meritocracy is also high, particularly when it comes to the reward of
effort, with an striking eighty six percent of agreement. Regarding
these descriptive indicators, a a first message coming from this study
is that there is something particular in school-age when it comes to
perceptions and preferences about inequality. This can relate to
development aspects
\citep[\citet{rizzo_children_2020}]{kim_socioeconomic_2020} but it could
be something specific to the Chilean case, which requires further
comparative as well as longitudinal research in the area. In any case,
teaching about the origins and characteristics of social inequality is
not part of the chilean school curriculum, calling for a more dedicated
consideration of these topics in areas such as citizensip education as
well as in history and social science education.

@@

The consideration of the school context opened some caveats in this
research. We found contrary evidence for the association between status,
school performance, and market justice preferences: high status and high
achievement at the school level are associated with less market justice
preferences. This is opposed to the rational-interest argument that
suggests that those doing better - both economically and academically-
would demand less redistribution \citep[\citet{runst_does_2018},
\citet{bullock_education_2021}]{gonthier_parallel_2017} and therefore
are expected to challenge market mechanisms. A possible explanation is
that education contributes to forming a critical view of markets in
redistribution, nevertheless, further research in this area is needed to
assert the specific mechanisms that might be playing a role. All in all,
from this study we know that school contexts are not trivial in this
regard, calling for larger attention to school-age research on the
formation of inequality preferences and beliefs.

@@

\dataavailability{The original data presented in the study are openly
available in
\url{https://www.agenciaeducacion.cl/orientar/estudios/estudios-nacionales/}}

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The socialization of meritocracy and market justice preferences at school},
  % pdfauthor={Juan Carlos Castillo; Mauricio Salgado; Kevin Carrasco; Andreas Laffert},
  pdfkeywords={market
justice, meritocracy, socialization, family, schools, Chile},
  pdfcreator={LaTeX via pandoc},
  unicode=true,
	bookmarksopen={true},
	pdffitwindow=true, 
	colorlinks=true, 
	linkcolor=bluecite, 
	citecolor=bluecite, 
	urlcolor=bluecite, 
	hyperfootnotes=true, 
	pdfstartview={FitH},
	pdfpagemode= UseNone}

@@

%\Author{Juan Carlos Castillo$^{1}$\orcidA{}, 
%Mauricio Salgado$^{2}$\orcidB{}, 
%Kevin Carrasco$^{3}$\orcidC{}~and~Andreas Laffert$^{4}$\orcidD{}
%}
%\longauthorlist{yes}

% MDPI internal command: Authors, for metadata in PDF
%\AuthorNames{
%  Juan Carlos Castillo,
%  Mauricio Salgado,
%  Kevin Carrasco,
%  Andreas Laffert}

% MDPI internal command: Authors, for citation in the left column
%\AuthorCitation{Lastname, F.; Lastname, F.; Lastname, F.}
%\usepackage{xstring}
%\newcommand\getfirst[1]{\getfirstaux#1\relax} \def\getfirstaux#1#2\relax{#1}
%\newcommand{\authorformatted}{Castillo, \getfirst{Juan Carlos}.; Salgado, \getfirst{Mauricio}.; Carrasco, \getfirst{Kevin}.; Laffert, \getfirst{Andreas}.}
%\AuthorCitation{\authorformatted}

@@

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
%\address {%
%$^{1}$ \quad Department of Sociology, Universidad de Chile\\
%$^{2}$ \quad CEP, Centro de Estudios Públicos\\
%$^{3}$ \quad COES, Center for Social Conflict and Cohesion Studies\\
%$^{4}$ \quad Institute of Sociology, Pontificia Universidad Católica de Chile\\
%}

@@

% Contact information of the corresponding author
%% see: https://quarto.org/docs/journals/authors.html
%\corres{Correspondence: juancastillov@uchile.cl}

@@

Robert E. Lane proposed the underpinnings of the concept of market
justice, which he differentiated from political justice. For him, ``it
is the genius of the market to stimulate wants without at the same time
stimulating a sense of deserving more than one gets''
\citep[p.384]{lane_market_1986}. Contrary to the evidence that
unequal distribution produces feelings of dissatisfaction, anger, and
resentment that might motivate forms of collective action
\citep{greitemeyer_subjective_2016, mishra_subjective_2015, smith_relative_2012, power_deprivationprotest_2018},
Lane pointed out that in market settings, social comparisons are more
likely to motivate increased effort rather than feelings of acute
injustice because individuals attribute outcomes to their actions. In
this sense, unequal levels of well-being would be, to some extent, a
function of their talents and efforts, instead of being based on
distributive principles that characterize welfare states, such as need
and equality (see \citep{wilson_role_2003}).

@@

Our research aimed to examine the relationship between the perception of
meritocracy and the justification of market justice preferences among
eighth-grade students in Chile. From our knowledge this is the first
study addressing market justice preferences at school level, finding
that students exhibit large preferences for market justice (about one
third agree or strongly agree with them), a share that is quite larger
when compared with evidence in adult population. Perception of
meritocracy is also high, particularly when it comes to the reward of
effort, with an striking eighty six percent of agreement. Regarding
these descriptive indicators, a a first message coming from this study
is that there is something particular in school-age when it comes to
perceptions and preferences about inequality. This can relate to
development aspects
\citep{rizzo_children_2020, kim_socioeconomic_2020} but it could
be something specific to the Chilean case, which requires further
comparative as well as longitudinal research in the area. In any case,
teaching about the origins and characteristics of social inequality is
not part of the chilean school curriculum, calling for a more dedicated
consideration of these topics in areas such as citizensip education as
well as in history and social science education.

@@

The consideration of the school context opened some caveats in this
research. We found contrary evidence for the association between status,
school performance, and market justice preferences: high status and high
achievement at the school level are associated with less market justice
preferences. This is opposed to the rational-interest argument that
suggests that those doing better - both economically and academically-
would demand less redistribution \citep{runst_does_2018,bullock_education_2021,gonthier_parallel_2017} and therefore
are expected to challenge market mechanisms. A possible explanation is
that education contributes to forming a critical view of markets in
redistribution, nevertheless, further research in this area is needed to
assert the specific mechanisms that might be playing a role. All in all,
from this study we know that school contexts are not trivial in this
regard, calling for larger attention to school-age research on the
formation of inequality preferences and beliefs.

@@

\dataavailability{The original data presented in the study are openly
available in
\url{https://www.agenciaeducacion.cl/orientar/estudios/estudios-nacionales/}

Access to the data, code, and analysis presented in the study are openly available in \url{https://github.com/educacion-meritocracia/market-justice-preferences-school}

### 2024-08-05 papelera/paper.qmd
Actualización paper anonimizado en .tex y orden de repo

---
title: "The socialization of meritocracy and market justice preferences at school"
format:
  hikmah-pdf: 
    indent: false
  hikmah-manuscript-pdf: 
    indent: false
abstract: |
  Previous research has shown that schools justify student performance differences based on meritocratic ideals. However, the link between school meritocracy and students’ justification of inequalities has received little attention. This paper examines how school and family socialization influence students’ beliefs about meritocracy and market-based justifications for access to health, pensions, and education in Chile. Using data from the First Study of Civic Education in Chile, we employ a multilevel modeling strategy to analyze 5,047 students and parents nested within 231 schools. Our findings reveal that a significant proportion of Chilean students agree with market justice principles, at a higher rate than adults. Most students endorse meritocratic views, particularly the reward of effort. Multilevel estimations show strong associations between meritocratic beliefs and market justice preferences: students who believe in the reward of effort and talent are more likely to justify inequalities based on financial capacity. At the school level, market justice preferences are higher in high-status schools but lower in schools with better achievement. Additionally, the conditional influence of meritocracy beliefs weakens in schools with higher socioeconomic status and performance. These results suggest that meritocratic beliefs and market justice preferences begin early and are influenced by the school environment.
keywords: 
  - market justice 
  - meritocracy
  - socialization
  - family
  - schools
  - Chile
bibliography: "input/bib/just-ineq-merit.bib"
csl: "input/bib/multidisciplinary-digital-publishing-institute.csl"
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
geometry: "left=2.54cm,right=2.54cm,top=2.54cm,bottom=2.54cm"
papersize: letter
colorlinks: true
fig-height: 4
fig-width: 7.5
header-includes:
  - \usepackage{threeparttable}
---

>>>>

---
title: "The socialization of meritocracy and market justice preferences at school"
format:
  mdpi-pdf:
    keep-tex: true
mdpi_journal: behavsci
article_type: article
article_status: submit
author:
  - name: Juan Carlos Castillo
    email: juancastillov@uchile.cl
    affiliations: 
        name: Department of Sociology, Universidad de Chile
        country: Chile
    attributes:
      corresponding: true
  - name: Mauricio Salgado
    email: msalgado@unab.cl
    affiliations: 
        name: Centro de Estudios Públicos, Chile
        country: Chile
  - name: Kevin Carrasco
    email: kevin.carrasco@ug.uchile.cl
    affiliations: 
        name: Center for Social Conflict and Cohesion Studies
        country: Chile
  - name: Andreas Laffert
    email: alaffertt@estudiante.uc.cl
    affiliations: 
        name: Institute of Sociology, Universidad Católica de Chile
        country: Chile
abstract: |
  Previous research has shown that schools justify student performance differences based on meritocratic ideals. However, the link between school meritocracy and students’ justification of inequalities has received little attention. This paper examines how school and family socialization influence students’ beliefs about meritocracy and market-based justifications for access to health, pensions, and education in Chile. Using data from the First Study of Civic Education in Chile, we employ a multilevel modeling strategy to analyze 5,047 students and parents nested within 231 schools. Our findings reveal that a significant proportion of Chilean students agree with market justice principles, at a higher rate than adults. Most students endorse meritocratic views, particularly the reward of effort. Multilevel estimations show strong associations between meritocratic beliefs and market justice preferences: students who believe in the reward of effort and talent are more likely to justify inequalities based on financial capacity. At the school level, market justice preferences are higher in high-status schools but lower in schools with better achievement. Additionally, the conditional influence of meritocracy beliefs weakens in schools with higher socioeconomic status and performance. These results suggest that meritocratic beliefs and market justice preferences begin early and are influenced by the school environment.
keywords: 
  - market justice 
  - meritocracy
  - socialization
  - family
  - schools
  - Chile
bibliography: "input/bib/just-ineq-merit.bib"
csl: "input/bib/multidisciplinary-digital-publishing-institute.csl"
header-includes:
  - \usepackage{threeparttable}
  - \usepackage{graphicx}
  - \includegraphics{pics/logo-mdpi.pdf}
---
